---
title: "When did your name peak in New Zealand?"
author: "Lee Durbin"
format: html
editor: visual
---

The table below contains data of each first name of babies born in Aotearoa New Zealand from 1900 to 2017. The data set includes all names with more than 10 instances in any given year.

This data was sourced from the [nzbabynames R package](https://github.com/ekothe/nzbabynames), and originally came from <https://catalogue.data.govt.nz/dataset/baby-name-popularity-over-time/>

```{r}
#| echo: false
#| warning: false
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(reactablefmtr) # Streamlined Table Styling and Formatting for Reactable
```

Search for your name to see when it last featured in the data, when and by how much it peaked, and how its popularity has changed over time.

```{r}
#| echo: false
names <- nzbabynames::nzbabynames |> 
  mutate(Name = stringi::stri_conv(Name, from = "latin1", to = "UTF-8"))

last_featured <- names |> 
  group_by(Name) |> 
  filter(Year == max(Year)) |> 
  ungroup() |> 
  select(Name, last_year_featured = Year, last_year_count = Count)

max_year <- names |> 
  group_by(Name) |> 
  filter(Count == max(Count)) |> 
  ungroup() |> 
  select(Name, max_year = Year, max_year_value = Count)

name_stats <- left_join(last_featured, max_year, by = "Name")

names_over_time <- names |> 
  left_join(name_stats, by = "Name") |> 
  with_groups(c(Name, last_year_featured, max_year, max_year_value), summarise, names_data = list(Count))
  
reactable(
  names_over_time,
  theme = fivethirtyeight(),
  fullWidth = FALSE,
  showPageInfo = FALSE,
  paginationType = "jump",
  defaultPageSize = 25,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(25, 50, 75, 100),
  filterable = TRUE,
    columns = list(
      Name = colDef(maxWidth = 200),
      last_year_featured = colDef(
        name = "Last Featured",
        maxWidth = 100,
        filterable = FALSE
        ),
      max_year = colDef(
        name = "Year Peaked",
        maxWidth = 100,
        filterable = FALSE
        ),
      max_year_value = colDef(
        name = "Peak Year Total",
        maxWidth = 100,
        filterable = FALSE
        ),
      names_data = colDef(
        cell = react_sparkline(
          names_over_time,
          highlight_points = highlight_points(max="blue")
          ),
        filterable = FALSE,
        name = "Trend",
        minWidth = 300
      )
    )
  )

```

